---
title: "Sleep Cycles"
author: "Mathieu C."
date: "1/17/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Data exploration using a 3d scatterplot

### Advantages

- Allows to see the shape of some distributions better than with a traditional scatterplot
- Can show up to 5 dimensions in a (somewhat) interpretable fashion (x,y,z,color,shape)
- Is fun to play with

## Description

### What we're looking at:
The data in the following slide comes from the `msleep` database, it is one of the databases included in R and accessible through the `data()`function.

### A glimpse into the database
Columns 1-5 from `msleep` are factors (name, genus, alimentation method, order and conservation status, respectively)

Columns 6-11 are numeric and represent different aspects of the sleep cycle of the observed animals. These are the ones we're going to use.  

## Data informations and transformations

Because we need values in all three dimensions to build the plot, we'll quickly remove unwanted rows containing `NA`'s using the `complete.case()` function beforehand, resulting in 30 rows of data and 11 variables.  
  
The values plotted are:

1. X : sleep_cycle which is the length of one sleep cycle in minutes (transformed from hours)
2. Y : sleep_total which is the ammount of sleep in minutes (transformed from hours)
3. Z : bodywt which is the body weight in $log_2(kilogramms)$ (transformed to reduce skew)

## The plot:

```{r cycles, echo=FALSE, warning=FALSE, comment="",message=FALSE}
require(plotly)
data("msleep")
cc <- complete.cases(msleep[,c("sleep_cycle", "sleep_total", "bodywt", "brainwt")])
df.cc <- msleep[cc,]

p <- plot_ly(data=df.cc, x=df.cc$sleep_cycle*60, y=df.cc$sleep_total*60,
             z=log(df.cc$bodywt,2), 
             type = "scatter3d",
             color = df.cc$brainwt*1000) %>%
        layout(title = "Animal Sleep Patterns\nColored by Brain Weight in gramms:",
               scene= list(
                       xaxis = list(title = "X: Sleep Cycle (minutes)"),
                       yaxis = list(title = "Y: Sleep Total (minutes)"),
                       zaxis = list(title ="Z: Body Weight (log2(kilogramms))")
                       )
               )
p
```
  
## Conclusions, Observations
With this 3d Scatterplot covering 4 dimensions, you can already observe trends between the observed variables like `sleep_total` and `bodywt` (cor = 0.60) or `brainwt` and `sleep_cycle` (cor = 0.85).  
  
`plot_ly()` is a fun and efficient way to create such a plot within minutes.

Thank you for your attention.
